Climate-driven Phytoplankton community variability in the Cariaco Basin, Venezuela

### 2024-12-19 agujournaltemplate.tex

    Between late 1995 and early 1998, nutrient data were supplied by William Senior (Universidad de Oriente, Cumana, Venezuela) and from mid-1997 to the present by Dr. K. Fanning (University of South Florida, St. Petersburg, FL). Samples from eight cruises were analyzed by both laboratories. Data for $PO_4$ and $SiO_4$ were in strong agreement (within ±5\%), whereas the $NO_x$ and $NH_4$ measurements showed discrepancies. \cite{taylor_ecosystem_2012}. For the overview of the entire time-series, we used a merged dataset for all nutrients, where coincidental monthly measurements were averaged.

    For the identification of phytoplankton species by microscopy, 500 mL seawater samples were collected at the same depths in HDPE bottles and preserved with a 5\% (final concentration) formalin solution neutralized with sodium tetraborate. Quantitative analysis was performed at the Universidad de Oriente, Venezuela, following the Utermöhl method \cite{hasle1978inverted}, using 100 mL sedimentation chambers with a 48-hour settling period. Phytoplankton identification and enumeration by microscopy were restricted to microphytoplankton (cells $>$20 μm) \cite{mutshinda_environmental_2013}.
    %from Mutshinda et al. 2013
    Quality control the microscopy data was performed by assembling the taxonomic identifications made across all cruises, removing observations that were not identified to species or genus level, correcting spelling variations, and grouping synonyms using taxonomic information in the World Register of Marine Species (\url{marinespecies.org}). After processing the data, 482 phytoplankton species were observed over the course of the time series, which were grouped into 223 phytoplankton genera.

    The datasets for Niskin bottle samples, CTD and phytoplankton taxonomy were retrieved from the Biological and Chemical Oceanography Data Management Office (BCO-DMO) \url{https://www.bco-dmo.org/project/2047} (accessed February, 2024). 
    From the CARIACO Time Series data, we used only those variables that cover the entire length of the time series without large gaps in coverage. Therefore, we did not include Zooplankton net tow and HPLC pigment data in our analysis.

@@

% Figure 3
Taking the microscopy community data, interpolated within the top 100 meters, we wanted to find if there was an apparent structure within the time series that could be related to the physical environment. For this, we used different clustering methods, both on the annual aggregate of cell counts from microscopy and on the individual monthly measurements. Figure \ref{fig:clustering} $a$ shows the results of clustering the annual aggregate communities with a linear complete linkage algorithm based on the binary Jaccard matrix. The community is separated into two groups, with the first cluster including the years 1996-2004 and 2014-2016 and the second cluster of the intermediate years from 2004-2013. This confirms previously observed patterns in the Cariaco ecosystem, where a strong shift in the community happened in 2004, but the community returned to the previous regime past 2014 \cite{taylor_ecosystem_2012, muller-karger_scientific_2019}.
Interannual species turnover is shown in Figure \ref{fig:clustering} $b$, where there is an increasing trend in turnover until 2004, where turnover is reduced for the period of cluster 2, but in 2014 with the return to cluster 1, turnover is higher again. This pattern is also reflected in the clustering in Figure \ref{fig:clustering} $a$, where cluster 2 shows a lower distance in the clustering tree between the years. 
To check whether annual aggregation of the community data smoothed out strong seasonal signals, we performed an NMDS cluster analysis using the monthly cell counts (see Figure \ref{fig:clustering} $c$). This NMDS clustering is overlayed with environmental data and climate indices to help explain some of the underlying trends. The first axis separates the two clusters relatively cleanly, where cluster 2 is trending towards postive AMO anomalies and cluster 2 trends towards positive anomalies in salinity and the MEI v.2 index. The second NMDS axis appears to follow the trends in upwelling intensity as described by the opposing trend arrows for nutrient concentrations and temperature, as well as wind speed. Overall, NMDS clustering seems to confirm the pattern found by the annual aggregate clustering. 
Figure \ref{fig:clustering} $d$ shows density distributions for environmental variables, climate and diversity indices between the two clusters. The patterns for upwelling related variables show clear differences between the two clusters, for example for cluster 1 there was a higher occurrence of high wind, shallower 21 \degree C Isotherm, as well as lower temperature and higher salinity and nutrient concentrations, in particular for nitrate. The AMO index shows a striking separation between the two clusters, where cluster 1 shows more negative anomalies, while cluster 2 has a high occurrence of positive anomalies. Although the mean diversity for all the indices is relatively similar, with a slightly higher mean in the Shannon and Pielou index for cluster 2, the density distribution shows marked differences. For cluster 1 there is a higher variance in genus richness, while cluster 2 shows higher variance and a bi-modal distribution for Shannon and Pielou indices.

@@

\begin{figure}
\label{fig:zscore}
\noindent\includegraphics[width=\textwidth]{fig/PLOTZScores_updated.pdf}
\caption{Heatmap of yearly average values, normalized for comparison as z-scores. All data is available at monthly resolution (see next draft figure for raw data). Red indicates a larger value, blue indicates a smaller value for that particular year. The years 1995 (n=2) and 2017 (n=1) were removed due to lack of data. Diversity Indices are calculated based on microscopy counts at the Genus level.}
\end{figure}

@@

\begin{figure}
\label{fig:clustering}
\noindent\includegraphics[width=\textwidth]{fig/ClusteringCompPlot_NEW.pdf}
\caption{(A) Clustering of microscopy data at the Genus level, using the binary Jaccard distance. The community for each year is binned together, to smooth out seasonal aspects. (B) NMDS plot of monthly community data (cell counts) that is cube root transformed. Stress: 0.261, 2 dimensions, converging solution. Color of dots based on which yearly cluster the monthly data is in. Vector overlay of ENVFIT shows underlying environmental variables that are co-varying with the community data.
Plots (A-L) below show density distributions of the environmental variables as well as diversity indices between the clustered years. Dashed line shows the mean.}
\end{figure}

>>>>

    Between late 1995 and early 1998, nutrient data was provided by William Senior (Universidad de Oriente, Cumana, Venezuela) and from mid-1997 to the present by Dr. K. Fanning (University of South Florida, St. Petersburg, FL). Samples from eight cruises were analyzed by both laboratories. Data for $PO_4$ and $SiO_4$ were in strong agreement (within ±5\%), whereas the $NO_x$ and $NH_4$ measurements showed discrepancies. \cite{taylor_ecosystem_2012}. For our analysis, we used a merged dataset for all nutrients, where coincidental monthly measurements were averaged.

    For the identification of phytoplankton species by microscopy, 500 mL seawater samples were collected at the same depths in HDPE bottles and preserved with a 5\% (final concentration) formalin solution neutralized with sodium tetraborate. Quantitative analysis was performed at the Universidad de Oriente, Venezuela, following the Utermöhl method \cite{hasle1978inverted}, using 100 mL sedimentation chambers with a 48-hour settling period. Phytoplankton identification and enumeration by microscopy were restricted to microphytoplankton (cells $>$20 μm) \cite{mutshinda_environmental_2013}.
    %from Mutshinda et al. 2013
    Quality control the microscopy data was performed by assembling the taxonomic identifications made across all cruises, removing observations that were not identified at the species or genus level, correcting spelling variations and grouping synonyms using taxonomic information from the World Register of Marine Species (\url{marinespecies.org}). After processing the data, 482 phytoplankton species were observed over the course of the time series, which were grouped into 223 phytoplankton genera.

    Data sets for Niskin bottle samples, CTD, and phytoplankton taxonomy were retrieved from the Biological and Chemical Oceanography Data Management Office (BCO-DMO) \url{https://www.bco-dmo.org/project/2047} (accessed February, 2024). 
    From the CARIACO Time Series data, we used only those variables that cover the entire length of the time series without large gaps in coverage. Therefore, we did not include Zooplankton net tow and HPLC pigment data in our analysis.

@@

% Figure 3
Taking the microscopy community data, interpolated within the top 100 meters, we wanted to find if there was an apparent structure within the time series that could be related to the physical environment. For this, we used different clustering methods, both on the annual aggregate of cell counts from microscopy and on the individual monthly measurements. Figure \ref{fig:clustering} $a$ shows the results of clustering the annual aggregate communities with a linear complete linkage algorithm based on the binary Jaccard matrix. The community is separated into two groups, with the first cluster including the years 1996-2004 and 2014-2016 and the second cluster of the intermediate years from 2004-2013. This confirms previously observed patterns in the Cariaco ecosystem, where a strong shift in the community happened in 2004, but the community returned to the previous regime past 2014 \cite{taylor_ecosystem_2012, muller-karger_scientific_2019}.
Interannual species turnover is shown in Figure \ref{fig:clustering} $b$, where there is an increasing trend in turnover until 2004, where turnover is reduced for the period of cluster 2, but in 2014 with the return to cluster 1, turnover is higher again. This pattern is also reflected in the clustering in Figure \ref{fig:clustering} $a$, where cluster 2 shows a lower distance in the clustering tree between the years. 
To check whether annual aggregation of the community data smoothed out strong seasonal signals, we performed an NMDS cluster analysis using the monthly cell counts (see Figure \ref{fig:clustering} $c$). This NMDS clustering is overlaid with environmental data and climate indices to help explain some of the underlying trends. The first axis separates the two clusters relatively cleanly, where cluster 2 is trending towards positive AMO anomalies and cluster 2 trends towards positive anomalies in salinity and the MEI v.2 index. The second NMDS axis appears to follow the trends in upwelling intensity as described by the opposing trend arrows for nutrient concentrations and temperature, as well as wind speed. Overall, NMDS clustering seems to confirm the pattern found by the annual aggregate clustering. 
Figure \ref{fig:clustering} $d$ to $o$ shows density distributions for environmental variables, climate and diversity indices between the two clusters. The patterns for upwelling related variables show clear differences between the two clusters, for example for cluster 1 there was a higher occurrence of high wind, shallower 21 \degree C Isotherm, as well as lower temperature and higher salinity and nutrient concentrations, in particular for nitrate. The AMO index shows a striking separation between the two clusters, where cluster 1 shows more negative anomalies, while cluster 2 has a high occurrence of positive anomalies. Although the mean diversity for all the indices is relatively similar, with a slightly higher mean in the Shannon and Pielou index for cluster 2, the density distribution shows marked differences. For cluster 1 there is a higher variance in genus richness, while cluster 2 shows higher variance and a bi-modal distribution for Shannon and Pielou indices.

@@

\begin{figure}
\label{fig:zscore}
\noindent\includegraphics[width=\textwidth]{fig/PLOTZScores_updated.pdf}
\caption{Heatmap of yearly average values, normalized for comparison as z-scores. Red indicates a larger value, blue indicates a smaller value for that particular year. The years 1995 (n=2) and 2017 (n=1) were removed due to lack of data, additionally we marked years with more than . Diversity Indices are calculated based on microscopy counts at the Genus level.}
\end{figure}

@@

\begin{figure}
\label{fig:clustering}
\noindent\includegraphics[width=\textwidth]{fig/ClusteringCompPlot_NEW 2.pdf}
\caption{(A) Clustering of microscopy data at the Genus level, using the binary Jaccard distance. The community for each year is binned together, to smooth out seasonal aspects. (B) NMDS plot of monthly community data (cell counts) that is cube root transformed. Stress: 0.261, 2 dimensions, converging solution. Color of dots based on which yearly cluster the monthly data is in. Vector overlay of ENVFIT shows underlying environmental variables that are co-varying with the community data.
Plots (A-L) below show density distributions of the environmental variables as well as diversity indices between the clustered years. Dashed line shows the mean.}
\end{figure}
